CENSE - Perceptual test II

### 2019-02-26 report.tex
Corrections and proof read

This report presents the results of a perceptual experiment conducted as part of the Cense project~\cite{cense}. From recordings of a sensor network with low information content, this study aims at proposing a processing chain for the estimation of the pleasantness in urban sound environments.

@@

With respect to the three discussed objectives, a test corpus is created based on recordings from one of the four soundwalks studied in~\cite{aumond}. The scenes correspond to 19 locations in the 13th district of Paris. For each recording 45 seconds of audio in a single channel are extracted that represent well the properties of their respective ambiances, and without single events overwhelming their overall perception.
The first objective requires the comparison of perceptual assessments for recorded and replicated sound scenes. The original recordings from 6 locations (P1, P3, P4, P8, P15 and P18) with diverse environments are picked to explore different real-life situations, as a function of the classifications proposed in~\cite{jrg}: \textit{parks}, \textit{quiet street}, \textit{noisy street} and \textit{very noisy street}. For the second part of the corpus sound scenes are simulated based on the 19 45~s extracts. The recordings are first annotated in terms of the following properties:

@@

The 19 scenes are then replicated using simScene~\cite{simscene} with a database of extracts from recordings of isolated sound sources. This database is constructed from part of the LibriSpeech~\cite{libri} corpus for voices and Freesound\footnote{\url{https://freesound.org}} contributions for all other sources. For this experiment, the playback sound level is the same as during recordings \ml{of the references scenes ?}, and ranges from 63.9~dB to 79.4~dB.\\

@@

To study the perceptual properties of simulated scenes, the corpus is extended with 75 simulated scenes from new scenarios. Because of the lack of large isolated sources datasets, we choose to restrict the scene composition to 3 sources: \textit{traffic}, \textit{human voices} and \textit{birds}. Considering the aim of the study, newly generated scenarios should cover most real-life situations while remaining perceptually plausible. Thus, statistics on background and event properties are extracted for each of the three sources from annotations presented in~\cite{jrg}. These statistics include, conditionally to the ambiance:
\begin{itemize}
\item The probability of appearance of a given source in the scene, for which events and backgrounds are considered separately,
\item Event-to-background ratios in dB for all events and backgrounds besides the first, in the form of normal distributions,
\item The inter-onset of event occurrences in seconds, also in the form of normal distributions.
\end{itemize}
Adjustments are made on the variance of the considered properties to better cover possible scenarios, and voice events EBR, mostly read English recordings, are reduced to improve realism. An additional \textit{place} ambiance is added with properties derived empirically from available data. For each environment diverse new scenarios are generated by sampling the background and event properties distributions. For each scene the time of presence for the three sources is estimated using indicators proposed in~\cite{gontier}, and 75 scenes are chosen to maximize pairwise distance in the resulting 3-dimensional space. This is shown in Figure~\ref{fig:tvb_pres}. The playback sound level is also conditioned on the ambiance according to Figure~\ref{fig:amb_levels}, and ranges from 46.6~dB to 77.1~dB.\\

@@

Test participants are asked to answer 9 questions per sound scene in the form of 11 point (0-10) scales. They are presented in french and translated in this report using standard terminology. The first 5 scales relate to general properties of the scene:
\begin{itemize}
\item Pleasantness (\textit{Unpleasant - Pleasant}) - P
\item Liveliness (\textit{Inert, amorphous - Lively, eventful}) - L
\item Overall loudness (\textit{Quiet - Noisy}) - OL
\item Interest (\textit{Boring, uninteresting - Stimulating, interesting}) - I
\item Calmness (\textit{Agitated, chaotic - Calm, peaceful}) - C
\end{itemize}
These quantities are typically studied in the perceptual characterization of sound scenes (CITE). While this work focuses on pleasantness, including other dimensions can be useful to verify the relevance of the data in relation to previous work.\\
Furthermore, the primary aim of this test is to model pleasantness from source-specific parameters. Previous studies (CITE) show that pleasantness is mainly influenced by the activity of three sources: traffic, human voices and birds. The perception of source activity can be linked to notions of dominance, time of presence,  or emergence. We choose the perceptual time of presence, that is the percentage of time for which the corresponding source is heard, as the activity descriptor for this study. The influence of traffic on pleasantness can be different for background and emergent events. Thus, 4 additional questions are presented:
\begin{itemize}
\item Sound level of passing vehicles (\textit{Very low - Very high}) - $L_{T, p}$,
\item Time of presence of Traffic, Voices, Birds (\textit{Never - Continuously}), resp. $T_{T, p}$, $T_{V, p}$ and $T_{B, p}$
\end{itemize}
The corresponding interface is shown in Figure~\ref{fig:interface}.

@@

Participants evaluate each scene on criteria coded on a discrete scales \ml{how many steps ?}. Many statistical tools assume that the studied data follows continuous, single mode distributions. It is thus important to first verify that no distinct groups of participants can be obtained from perceptual assessments. A hierarchical clustering analysis of the participants is performed for the 25 scenes in the commonly evaluated sub-corpus, and shown in Figure~\ref{fig:hclusters}. No clear separation into groups is found, as a result the assessment distributions are assumed to be single-mode and their means can be used for further analysis.

@@

Table~\ref{tab:ogrep} shows the mean differences between assessments for the recorded and corresponding replicated scenes. Kolmogorov-Smirnov statistical tests are implemented for each scene and perceptual scale to outline significant differences between assessment distributions, which are shown in bold. This non parametric test is chosen because it assumes that distributions are continuous. Pleasantness is found to differ between the two groups only for one location, P4, which can be explained by high discrepancies in the perceptual time of presence of traffic and voices. For this location, the background traffic in the recorded scene varies along time, it is louder in the first half of the scene than in the second half. Replicating this scene using simScene imposes a constant sound level for background sources. Thus, the background traffic is louder in the replicated scene than it is in the recording for about half
of its duration. To a lesser extent the same issue explains the large difference in assessed time of presence of voices in the location P3, \ml{trop long : but as voices have a less distinct influence on pleasantness this is not reflected directly}. Differences can also be interpreted by the choice of isolated samples, which is semi-random and based on a high-level source taxonomy. For example, no difference is made during annotation between child or adult speech, or depending on its expressiveness.\\

Overall, most perceptual responses for replicated scenes are close to those for recordings. No major difference is found in a particular scale, and the proposed hypothesis, \ml{that is that ...} cannot be invalidated.

@@

The perceptual space generated by the five general scales can also be studied and compared to previous studies to further validate the test material. It is obtained by performing a principal components analysis (PCA) on the corresponding perceptual responses at the scene level. Figure~\ref{fig:pspace} compares the results for scenes based on recordings (left, n=25) and new scenarios (right, n=75). The generated spaces are similar, with only overall loudness and pleasantness axes slightly rotated between the two corpora. For both sets the variance explained by the first two components is similar (resp. 79.4\% - 18.1\% and 79.6\% - 15.2\%). Furthermore, these representations are comparable to those found in previous work on perceptual dimensions~\cite{axelsson}. Thus, the use of simulated scenes based on both real and new scenarios does not result in major differences in the relations between perceptual quantities. This confirms that the simulation process generates perceptually convincing scenes and can be used for further research.\\

@@

We then study the relations between perceptual scales with respect to existing work. Table~\ref{tab:percc} shows the Pearson's correlation coefficients at the scene level, that is on the mean responses for each scene. The resulting values are consistent with the literature~\cite{aumond, lavandier}, both between general scales as discussed in the previous section and for the relations to source contributions. Pleasantness (P) is mainly influenced negatively by overall loudness and traffic and positively by birds. In previous studies a small positive contribution of voices to pleasantness was found, while no direct relation is visible from the data gathered. This can be explained by the choice of speech samples used during generation, which consist of read audiobooks extracts and thus may sound unnatural in the considered urban environments. Liveliness (L) is here negatively though weakly (r<0.4) correlated with the presence of birds, while no significant relation was found in previous studies.\\

@@

Table~\ref{tab:percm} summarizes the performance of the three models. Again, the two models computed on the present corpus display similar, low estimation errors as well as no significant difference in adjusted $R^2$ and correlation to assessed pleasantness. In contrast, the model proposed in~\cite{ricciardi2014} results in higher prediction errors, along with less data variance explained by the model. However, this model was developed on a large corpus comprised of more than 3400 sound scenes and still shows acceptable performance for this new corpus.

@@

Several indicators are computed from the audio signals that are found in previous studies~\cite{aumond, gontier, lavandier} to correlate well with content-related perceptual parameters: the overall loudness and the source-specific time of presence.\\

First, indicators can be computed directly from the mixed audio, without the need for information about the composition of the scene, \textit{i.e.} when the source are active and at which level. Typically this includes indicators derived from sound level measurements. For this study the following are computed with 125~ms (fast) measurements using the Matlab ITA-toolbox~\cite{ita}:

\begin{itemize}
\item Z-weighted $L_{eq}$ and A-weighted $LA_{eq}$ equivalent sound levels in dB and dBA respectively.
\item $L_{10}$, $L_{50}$ and $L_{90}$: 10th, 50th and 90th percentiles of the Z-weighted sound level. The $L_{10}$ is associated to events and the $L_{90}$ to background activity, while the $L_{50}$ is a measurement of the overall sound level.
\item $LA_{50}$: 50th percentile of the A-weighted sound level, with similar properties as the $L_{50}$.
\item $L_{50, 1kHz}$: 50th percentile of the Z-weighted sound level for the 1kHz frequency band, also a good descriptor of the overall sound level of the scene.
\item The time and frequency second derivative for the 4kHz band ($TFSD_{4kHz}$) introduced by~\cite{aumond}, that is found to correlate well with the perceptual time of presence of birds.
\end{itemize}

@@

Next, the $\hat T_s(\alpha)$ and $\hat T_s(\alpha, \beta)$ time of presence approximations proposed in~\cite{gontier} are computed. $\hat T_s(\alpha)$ is based on a binary emergence model using the full-band source emergence $\Delta_s(t) = L_s(t) - L_{\bar{s}}(t)$ and a threshold $\alpha$ optimized on the present dataset:

@@

Table~\ref{tab:physc} shows the Pearson's correlation coefficients between computed physical indicators and perceptual assessments. First, all measurements of global sound levels correlate well (r>0.9) with the perceived overall loudness. Although any could be used, the $L_{50}$ appears to be best suited for pleasantness prediction in this study \ml{because ??}.\\

@@

@@

Predicting these quantities directly \ml{from what ?} using deep learning architectures is difficult: training this type of models requires large annotated datasets which would require extensive work by several humans to obtain relevant labels. However, the perceived time of presence can be approximated by a simple masking model as demonstrated in Section~\ref{sec:physm}. While the same need for annotation of large datasets exists, it can be circumvented by the use of simulated data, for which the contribution of each source in the mix is available. Thus, a deep learning model can be used to predict either of the first steps of the pleasantness estimation chain.\\

@@

In this report, two models for source-specific sound level and presence estimation are implemented and compared. \ml{manque un lien entre les deux derniers par}

\subsection{Data}

@@

The sensors implemented as part of the Cense project provide fast measurements of sound levels for each third-octave band in the $20~Hz - 12.5~kHz$ range in their normal operation. Thus, the model should take as input the same representation. Third-octave spectra are computed for 125~ms frames and 29 frequency bands of the mixed sound scenes. Inputs are then obtained by splitting the resulting spectrograms into frames with a 1~s duration. This scale is considered relevant for perception and was already used to compute the masking models proposed in the perceptual experiment results analysis. The input frames of dimension 29x8 are processed independently in this study.\\ \ml{cite sensor paper}

@@

\subsection{Model}

@@

@@

Models are optimized using the Adam algorithm~\cite{adam} on mini-batches of 64 1s examples. For all experiments a constant learning rate of $\lambda = 0.001$ is used. Training is performed during 10 epochs, that is iterations over the whole training set. As shown in Figure~\ref{fig:mdl_loss}, at that point the training loss is still decreasing but not the validation loss, indicating that the model starts overfitting the training data.

@@

Table~\ref{tab:perf_cmp} displays the performance of the two studied models for metrics at the physical level. The first model outputs very poor estimations of the source-specific sound level, but the resulting source presence accuracy is high at 92.51\%. As a result, the time of presence estimates are precise with a root mean-squared error of 0.12 on a 0-1 scale. This behavior may be explained by the fact that $\hat T_s(\alpha)$ indicator is too permissive: if some sources are close in terms of full-band sound level, the threshold $\alpha = -19dB$ allows rougher estimates for binary presence values. Thus, when using $\hat T_s(\alpha)$, it is sufficient to obtain good presence detection that the model output either very low values when the source is absent or poor estimates of the level when it is present. Here, the model fails at the source separation and level estimation task and is probably closer to a detection model.\\ \ml{Clarifier les deux derineire phrases}

@@

Subsequently, a deep learning model with state-of-the-art architecture is developed for this task. The model takes as input the fast third-octave band representation of the mixed scene and is trained on a corpus of simulated scenes to output presence labels in short time frames of 1s. Low pleasantness prediction errors are achieved on the perceptual experiment corpus: around 1.15 root mean-squared error on a 10-point scale which is found satisfying \ml{est que l'on peut comparer a la variance de jugement perceptif ?}.\\

Future work will focus on the study of the robustness of developed models to new environment and data. Here only three sources were considered during training, and a limited isolated sound sources database have been considered. Thus, pleasantness estimation chain needs to be evaluated using data gathered in the experimentation site, where the Cense sensor network is to be implemented. %, as it contains specific sound environments as well as unique sources that were not considered in this study.

\textbf{TODO}

\textbf{Text}

\begin{itemize}
\item Appendix A
\end{itemize}

\textbf{Figures}

\begin{itemize}
\item Pleasantness prediction chain
\item Deep learning architecture
\item Headphones calibration
\end{itemize}

@@

\begin{thebibliography}{9}
%\bibitem{axelsson2010}
%Axelsson, \"O.; Nilsson, M.E.; Berglund, B.
%A principal components model of soundscape perception. \textit{J. Ac. Soc. Am.}, 2010
%
%\bibitem{jeon2018}
%Jeon, J.Y.; Hong, J.Y.; Lavandier, C; Lafon, J.; Axelsson, \"O.; Hurtig, M.
%A cross-national comparison in assessment of urban park soundscapes in France, Korea, and Sweden through laboratory experiments. \textit{Applied Acoustics}, 2018
%
%
%\bibitem{aumond}
%Aumond, P.; Can, A.; De Coensel, B.; Botteldooren, D.; Ribeiro, C.; Lavandier, C.
%Modeling soundscape pleasantness using perceptual assessments and acoustic measurements along paths in urban context. \textit{Acta Acustica united with Acustica}, 2017
%
%\bibitem{ita}
%Berzborn, M.; Bomhardt, R.; Klein, J.; Richter, J.-G.; Vorlander, M.
%The ITA-Toolbox: An Open Source MATLAB Toolbox for Acoustic Measurements and Signal Processing. \textit{In proceedings of the 43th Annual German Congress on Acoustics}, 2017
%
%\bibitem{simscene}
%Rossignol, M.; Lafay, G.; Lagrange, M.; Misdariis, N.
%SimScene : a web-based acoustic scenes simulator, \textit{In proceedings of the 1st Web Audio Conference (WAC)}, 2015
%
%\bibitem{gloaguen}
%Gloaguen, J.-R.; Can, A.; Lagrange, M.; Petiot, J.-F.
%Creation of a corpus of realistic urban sound scenes with controlled acoustic properties. \textit{In proceedings of the 173rd Meeting of the Acoustical Society of America and the 8th Forum Acusticum (Acoustics'17)}, 2017
\end{thebibliography}

>>>>

This report presents the results of a perceptual experiment conducted as part of the Cense project~\footnote{\url{http://cense.ifsttar.fr/}}. From recordings of a sensor network with low information content, this study aims at proposing a processing chain for the estimation of the pleasantness in urban sound environments.

@@

With respect to the three discussed objectives, a test corpus is created based on recordings from one of the four soundwalks studied in~\cite{aumond2017}. The scenes correspond to 19 locations in the 13th district of Paris. For each recording 45 seconds of audio in a single channel are extracted that represent well the properties of their respective ambiances, and without single events overwhelming their overall perception.
The first objective requires the comparison of perceptual assessments for recorded and replicated sound scenes. The original recordings from 6 locations (P1, P3, P4, P8, P15 and P18) with diverse environments are picked to explore different real-life situations, as a function of the classifications proposed in~\cite{gloaguen2017}: \textit{parks}, \textit{quiet street}, \textit{noisy street} and \textit{very noisy street}. For the second part of the corpus sound scenes are simulated based on the 19 45~s extracts. The recordings are first annotated in terms of the following properties:

@@

The 19 scenes are then replicated using simScene~\cite{rossignol2015} with a database of extracts from recordings of isolated sound sources. This database is constructed from part of the LibriSpeech~\cite{panayotov2015} corpus for voices and Freesound\footnote{\url{https://freesound.org}} contributions for all other sources. For this experiment, the playback sound level is the same as during recordings of the references scenes, and ranges from 63.9~dB to 79.4~dB.\\

@@

To study the perceptual properties of simulated scenes, the corpus is extended with 75 simulated scenes from new scenarios. Because of the lack of large isolated sources datasets, we choose to restrict the scene composition to 3 sources: \textit{traffic}, \textit{human voices} and \textit{birds}. Considering the aim of the study, newly generated scenarios should cover most real-life situations while remaining perceptually plausible. Thus, statistics on background and event properties are extracted for each of the three sources from annotations presented in~\cite{gloaguen2017}. These statistics include, conditionally to the ambiance:
\begin{itemize}
\item The probability of appearance of a given source in the scene, for which events and backgrounds are considered separately,
\item Event-to-background ratios in dB for all events and backgrounds besides the first, in the form of normal distributions,
\item The inter-onset of event occurrences in seconds, also in the form of normal distributions.
\end{itemize}
Adjustments are made on the variance of the considered properties to better cover possible scenarios, and voice events EBR, mostly read English recordings, are reduced to improve realism. An additional \textit{place} ambiance is added with properties derived empirically from available data. For each environment diverse new scenarios are generated by sampling the background and event properties distributions. For each scene the time of presence for the three sources is estimated using indicators proposed in~\cite{gontier2018}, and 75 scenes are chosen to maximize pairwise distance in the resulting 3-dimensional space. This is shown in Figure~\ref{fig:tvb_pres}. The playback sound level is also conditioned on the ambiance according to Figure~\ref{fig:amb_levels}, and ranges from 46.6~dB to 77.1~dB.\\

@@

Test participants are asked to answer 9 questions per sound scene in the form of 11 point (0-10) scales. They are presented in french and translated in this report using standard terminology. The first 5 scales relate to general properties of the scene:
\begin{itemize}
\item Pleasantness (\textit{Unpleasant - Pleasant}) - P
\item Liveliness (\textit{Inert, amorphous - Lively, eventful}) - L
\item Overall loudness (\textit{Quiet - Noisy}) - OL
\item Interest (\textit{Boring, uninteresting - Stimulating, interesting}) - I
\item Calmness (\textit{Agitated, chaotic - Calm, peaceful}) - C
\end{itemize}
These quantities are typically studied in the perceptual characterization of sound scenes~\cite{axelsson2010, aumond2017, nilsson2007}. While this work focuses on pleasantness, including other dimensions can be useful to verify the relevance of the data in relation to previous work.\\
Furthermore, the primary aim of this test is to model pleasantness from source-specific parameters. Previous studies show that pleasantness is mainly influenced by the activity of three sources: traffic, human voices and birds. The perception of source activity can be linked to notions of dominance, time of presence,  or emergence. We choose the perceptual time of presence, that is the percentage of time for which the corresponding source is heard, as the activity descriptor for this study. The influence of traffic on pleasantness can be different for background and emergent events. Thus, 4 additional questions are presented:
\begin{itemize}
\item Sound level of passing vehicles (\textit{Very low - Very high}) - $L_{T, p}$,
\item Time of presence of Traffic, Voices, Birds (\textit{Never - Continuously}), resp. $T_{T, p}$, $T_{V, p}$ and $T_{B, p}$
\end{itemize}
The corresponding interface is shown in Figure~\ref{fig:interface}.

@@

Participants evaluate each scene on criteria coded on 10-point discrete scales. Many statistical tools assume that the studied data follows continuous, single mode distributions. It is thus important to first verify that no distinct groups of participants can be obtained from perceptual assessments. A hierarchical clustering analysis of the participants is performed for the 25 scenes in the commonly evaluated sub-corpus, and shown in Figure~\ref{fig:hclusters}. No clear separation into groups is found, as a result the assessment distributions are assumed to be single-mode and their means can be used for further analysis.

@@

Table~\ref{tab:ogrep} shows the mean differences between assessments for the recorded and corresponding replicated scenes. Kolmogorov-Smirnov statistical tests are implemented for each scene and perceptual scale to outline significant differences between assessment distributions, which are shown in bold. This non parametric test is chosen because it assumes that distributions are continuous. Pleasantness is found to differ between the two groups only for one location, P4, which can be explained by high discrepancies in the perceptual time of presence of traffic and voices. For this location, the background traffic in the recorded scene varies along time, it is louder in the first half of the scene than in the second half. Replicating this scene using simScene imposes a constant sound level for background sources. Thus, the background traffic is louder in the replicated scene than it is in the recording for about half
of its duration. To a lesser extent the same issue explains the large difference in assessed time of presence of voices in the location P3, but as voices have a less distinct influence on pleasantness this is not reflected directly. Differences can also be interpreted by the choice of isolated samples, which is semi-random and based on a high-level source taxonomy. For example, no difference is made during annotation between child or adult speech, or depending on its expressiveness.\\

Overall, most perceptual responses for replicated scenes are close to those for recordings. No major difference is found in a particular scale, and the proposed hypothesis that recorded and replicated scenes generate the same perceptual responses cannot be invalidated.

@@

The perceptual space generated by the five general scales can also be studied and compared to previous studies to further validate the test material. It is obtained by performing a principal components analysis (PCA) on the corresponding perceptual responses at the scene level. Figure~\ref{fig:pspace} compares the results for scenes based on recordings (left, n=25) and new scenarios (right, n=75). The generated spaces are similar, with only overall loudness and pleasantness axes slightly rotated between the two corpora. For both sets the variance explained by the first two components is similar (resp. 79.4\% - 18.1\% and 79.6\% - 15.2\%). Furthermore, these representations are comparable to those found in previous work on perceptual dimensions~\cite{axelsson2010, cain2013}. Thus, the use of simulated scenes based on both real and new scenarios does not result in major differences in the relations between perceptual quantities. This confirms that the simulation process generates perceptually convincing scenes and can be used for further research.\\

@@

We then study the relations between perceptual scales with respect to existing work. Table~\ref{tab:percc} shows the Pearson's correlation coefficients at the scene level, that is on the mean responses for each scene. The resulting values are consistent with the literature~\cite{aumond2017, gontier2018}, both between general scales as discussed in the previous section and for the relations to source contributions. Pleasantness (P) is mainly influenced negatively by overall loudness and traffic and positively by birds. In previous studies a small positive contribution of voices to pleasantness was found, while no direct relation is visible from the data gathered. This can be explained by the choice of speech samples used during generation, which consist of read audiobooks extracts and thus may sound unnatural in the considered urban environments. Liveliness (L) is here negatively though weakly (r<0.4) correlated with the presence of birds, while no significant relation was found in previous studies.\\

@@

Table~\ref{tab:percm} summarizes the performance of the three models. Again, the two models computed on the present corpus display similar, low estimation errors as well as no significant difference in adjusted $R^2$ and correlation to assessed pleasantness. In contrast, the model proposed in~\cite{ricciardi2014} results in higher prediction errors, along with less data variance explained by the model. However, this model was developed on a large corpus comprised of more than 3400 sound scenes and still shows acceptable performance for this new corpus: the mean standard deviation of pleasantness assessments for this experiment is about 1.77.

@@

Several indicators are computed from the audio signals that are found in previous studies~\cite{aumond2017, gontier2018} to correlate well with content-related perceptual parameters: the overall loudness and the source-specific time of presence.\\

First, indicators can be computed directly from the mixed audio, without the need for information about the composition of the scene, \textit{i.e.} when the source are active and at which level. Typically this includes indicators derived from sound level measurements. For this study the following are computed with 125~ms (fast) measurements using the Matlab ITA-toolbox~\cite{itatoolbox2017}:

\begin{itemize}
\item Z-weighted $L_{eq}$ and A-weighted $LA_{eq}$ equivalent sound levels in dB and dBA respectively.
\item $L_{10}$, $L_{50}$ and $L_{90}$: 10th, 50th and 90th percentiles of the Z-weighted sound level. The $L_{10}$ is associated to events and the $L_{90}$ to background activity, while the $L_{50}$ is a measurement of the overall sound level.
\item $LA_{50}$: 50th percentile of the A-weighted sound level, with similar properties as the $L_{50}$.
\item $L_{50, 1kHz}$: 50th percentile of the Z-weighted sound level for the 1kHz frequency band, also a good descriptor of the overall sound level of the scene.
\item The time and frequency second derivative for the 4kHz band ($TFSD_{4kHz}$) introduced by~\cite{aumond2017}, that is found to correlate well with the perceptual time of presence of birds.
\end{itemize}

@@

Next, the $\hat T_s(\alpha)$ and $\hat T_s(\alpha, \beta)$ time of presence approximations proposed in~\cite{gontier2018} are computed. $\hat T_s(\alpha)$ is based on a binary emergence model using the full-band source emergence $\Delta_s(t) = L_s(t) - L_{\bar{s}}(t)$ and a threshold $\alpha$ optimized on the present dataset:

@@

Table~\ref{tab:physc} shows the Pearson's correlation coefficients between computed physical indicators and perceptual assessments. First, all measurements of global sound levels correlate well (r>0.9) with the perceived overall loudness. Although any could be used, the $L_{50}$ displays the highest correlation to pleasanntness and is thus selected for pleasantness prediction in this study.\\

@@

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.8\textwidth]{figures/ppred.pdf}
    \caption{Processing chain applied to obtain an estimation of pleasantness in a recorded scene.}\label{fig:ppred}
\end{figure}

@@

Predicting these quantities directly from audio recordings using deep learning architectures is difficult: training this type of models requires large annotated datasets which would require extensive work by several humans to obtain relevant labels. However, the perceived time of presence can be approximated by a simple masking model as demonstrated in Section~\ref{sec:physm}. While the same need for annotation of large datasets exists, it can be circumvented by the use of simulated data, for which the contribution of each source in the mix is available. Thus, a deep learning model can be used to predict either of the first steps of the pleasantness estimation chain.\\

@@

In the following sections, two models for source-specific sound level and presence estimation are implemented and compared.

\subsection{Data}
\label{sec:pred_data}

@@

The sensors implemented as part of the Cense project provide fast measurements of sound levels for each third-octave band in the $20~Hz - 12.5~kHz$ range in their normal operation~\cite{gontier2017}. Thus, the model should take as input the same representation. Third-octave spectra are computed for 125~ms frames and 29 frequency bands of the mixed sound scenes. Inputs are then obtained by splitting the resulting spectrograms into frames with a 1~s duration. This scale is considered relevant for perception and was already used to compute the masking models proposed in the perceptual experiment results analysis. The input frames of dimension 29x8 are processed independently in this study.\\

@@

\subsection{Model}
\label{sec:pred_mdl}

@@

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.8\textwidth]{figures/deep_arch.pdf}
    \caption{Architecture of deep learning models used for sound level or source presence prediction.}\label{fig:deep_arch}
\end{figure}

@@

Models are optimized using the Adam algorithm~\cite{kingma2015} on mini-batches of 64 1s examples. For all experiments a constant learning rate of $\lambda = 0.001$ is used. Training is performed during 10 epochs, that is iterations over the whole training set. As shown in Figure~\ref{fig:mdl_loss}, at that point the training loss is still decreasing but not the validation loss, indicating that the model starts overfitting the training data.

@@

Table~\ref{tab:perf_cmp} displays the performance of the two studied models for metrics at the physical level. The first model outputs very poor estimations of the source-specific sound level, but the resulting source presence accuracy is high at 92.51\%. As a result, the time of presence estimates are precise with a root mean square error of 0.12 on a 0-1 scale. This behavior may be explained by the fact that $\hat T_s(\alpha)$ indicator is too permissive: if some sources are close in terms of full-band sound level, the threshold $\alpha = -19dB$ allows rougher estimates for binary presence values. Thus, because the sound level model output is used in $\hat T_s(\alpha)$, there are two cases for which the correct presence is obtained
\begin{enumerate}
\item The model outputs very low level values for sources that are not present
\item The model outputs poor estimates but with sufficiently close values for all present sources
\end{enumerate}
Thus, the model may fail at the source separation and level estimation task and is here probably closer to a detection model.\\

@@

Subsequently, a deep learning model with state-of-the-art architecture is developed for this task. The model takes as input the fast third-octave band representation of the mixed scene and is trained on a corpus of simulated scenes to output presence labels in short time frames of 1s. Low pleasantness prediction errors are achieved on the perceptual experiment corpus: around 1.15 root mean square error on the 10-point scale which is below the mean standard deviation of pleasantness assessments of 1.77.\\

Future work will focus on the study of the robustness of developed models to new environment and data. Here only three sources were considered during training, and a limited isolated sound sources database have been considered. Thus, pleasantness estimation chain needs to be evaluated using data gathered in the experimentation site, where the Cense sensor network is to be implemented.

@@

\bibliographystyle{plain}
\bibliography{refs}
